
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(brms)
library(dplyr)
library(ggplot2)
library(lme4)
```


```{r, echo=FALSE}
dat <- read.csv("dat_1_.txt","", header = TRUE)
```


First six rows of the data-set are shown below
```{r, echo=FALSE}
head(dat)
```
# Defining a hierarchical linear model

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
#i first just want to try the frequentist version of this model
#by removing results='hide' & echo=FALSE the results and the code will be shown in the pdf document
m<-lmer(rt~1+so+(1+so|subj)+(1+so|item),dat)
summary(m)
```
#### defining a hierarchical linear model with varying intercepts and varying slopes for subject and item, assuming a correlation between the varying intercepts and slopes for both subject and item, and a gaussian likelihood.


```{r, message=FALSE, warning=FALSE, results='hide'}
m_hier_gaussian_brm<-brm(rt~so+(so|subj)+(so|item),
                data=dat,
                family = gaussian(),
                prior = 
                  c(prior(normal(0, 1000), class = Intercept),
                    prior(normal(0, 1000), class = b),
                    prior(normal(0, 500), class = sigma),
                    prior(normal(0, 500), class = sd),
                    prior(lkj(2), class = cor)),
                chains = 4)
```
```{r, echo=FALSE, results='hide'}
#the summary of the posteriors:
summary(m_hier_gaussian_brm)
```

summarize the posteriors of all parameters in a table (the fixed effect intercept and slope, the variance components, and the correlations)

```{r, echo=FALSE}
posterior_summary(m_hier_gaussian_brm)[1:9,c("Estimate","Q2.5","Q97.5")]
```

plotting the posterior distributions to obtain a graphical summary of all the parameters in the model

```{r, echo=FALSE}
plot(m_hier_gaussian_brm)
```

Histograms of eleven samples from the posterior predictive distribution of the model m_hier_gaussian_brm. the real data is skewed and has no values less than 0 ms while the predictive distributions are centered and symmetrical and have values less than 0 milliseconds (in order to evaluate our model)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
pp_check(m_hier_gaussian_brm, nsamples = 11, type = "hist")
```

posterior predictive checks with 100 predicted data-sets (in order to evaluate our model)

```{r, echo=FALSE}
#pp_check(m_hier_brm, nsamples = 100, type = "dens_overlay")

for (l in c(0.5,-0.5)){
df_dat <- filter(dat, so == l)
p <- pp_check(m_hier_gaussian_brm, type = "dens_overlay",
nsamples = 100,
newdata = df_dat) +
geom_point(data =df_dat, aes(x = rt, y = 0.0001))+
ggtitle(paste("so: ",l))
print(p)
}
```


#### defining a hierarchical linear model with varying intercepts and varying slopes for subject and item, assuming a correlation between the varying intercepts and slopes for both subject and item, and a lognormal likelihood.


```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
df_spacebar_ref <- dat %>%
   mutate(rt = runif(n(), 0, 10000))
fit_prior_press_trial <- brm(rt~so+(so|subj)+(so|item),
data = df_spacebar_ref,
family = lognormal(),
prior =  c(prior(normal(6, 1.5), class = Intercept),
                     prior(normal(0, 0.1), class = b),
                     prior(normal(0, 1), class = sigma),
                     prior(normal(0, 1), class = sd),
                     prior(lkj(2), class = cor)),
sample_prior = "only",
control = list(adapt_delta = .9)
)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
median_effect <- function(x){
median(x - lag(x), na.rm = TRUE)
}
pp_check(fit_prior_press_trial, type = "stat", stat = "median_effect")
```


```{r, message=FALSE, warning=FALSE, results='hide'}
m_hier_log_brm<-brm(rt~so+(so|subj)+(so|item),dat,
                 family = lognormal(),
                 prior =
                   c(prior(normal(6, 3), class = Intercept),
                     prior(normal(0, 0.1), class = b),
                     prior(normal(0, 1), class = sigma),
                     prior(normal(0, 1), class = sd),
                     prior(lkj(2), class = cor)),
                 iter = 4000,
                 chains = 4)
```

```{r, echo=FALSE, results='hide'}
#the summary of the posteriors:
summary(m_hier_log_brm)
```

summarize the posteriors of all parameters in a table (the fixed effect intercept and slope, the variance components, and the correlations)

```{r, echo=FALSE}
posterior_summary(m_hier_log_brm)[1:9,c("Estimate","Q2.5","Q97.5")]
```

plotting The posterior distributions to obtain a graphical summary of all the parameters in the model

```{r}
plot(m_hier_log_brm)
```
Histograms of eleven samples from the posterior predictive distribution of the model m_hier_brm1 (the real data is skewed and has no values less than 0 ms), also the predicted data.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
pp_check(m_hier_log_brm, nsamples = 11, type = "hist")
```
posterior predictive checks with 100 predicted data-sets (in order to evaluate our model)

```{r, echo=FALSE}
for (l in c(0.5,-0.5)){
df_dat <- filter(dat, so == l)
p <- pp_check(m_hier_log_brm, type = "dens_overlay",
nsamples = 100,
newdata = df_dat) +
geom_point(data =df_dat, aes(x = rt, y = 0.0001))+
ggtitle(paste("so: ",l))
print(p)
}
```